---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---
# Installations
```{r}
library(devtools)
devtools::install_github("JEFworks-Lab/MERINGUE")
BiocManager::install("BiocNeighbors")
devtools::install_github("KlugerLab/ALRA")
devtools::install_github("zdebruine/RcppML")
devtools:install_github("presto")

install.packages('NMF')
devtools::install_github("jokergoo/circlize")
devtools::install_github("jokergoo/ComplexHeatmap")

install.packages("Rfast", type = "source") 
# in case of error regarding Rfast
# when running CellChat Inference pathway

devtools::install_github("jinworks/SpatialCellChat")

```

# Load dependencies
```{r}
library(Seurat)
library(edgeR)
library(Matrix)
library(clusterProfiler)
library(org.Mm.eg.db)
library(enrichplot)
library(ggplot2)
library(ggrepel)
library(tidyverse)
library(presto)
library(conflicted)
library(readxl)

# CellChat specific dependencies
library(SpatialCellChat)
library(Matrix)
library(patchwork)
options(stringsAsFactors = F)
conflicted::conflicts_prefer(base::union)

source("scripts/helpers/project_paths_v3.R")
source("scripts/helpers/wholekidney_celltype_functions_v2.R")
source("scripts/helpers/plotting_functions_v3.R")
source("scripts/helpers/output_functions_v3.R")
source("scripts/helpers/pseudobulk_functions_v3.R")
```

# Load object
```{r}
xen1diet <- readRDS("D:/USERS/ELLIOT/Dissertation/xen1/xen1diet.rds")
```

# Add broad cell types
```{r}
xen1diet <- add_broad_celltypes(xen1diet)
```

```{r}
set.seed(1)

cells <- sample(colnames(xen1diet), 20000)

test_obj <- subset(xen1diet, cells = cells)
```

# Load ligand receptor pairs
```{r}
lr_raw <- read_xlsx("info/Xenium_Probe_List_May2025 3.xlsx", sheet = 2)

lr_pairs <- lr_raw[
  nzchar(trimws(lr_raw$Ligand)) & nzchar(trimws(lr_raw$Receptor)),
  c("Ligand", "Receptor")
]

lr_pairs$Ligand <- trimws(lr_pairs$Ligand)
lr_pairs$Receptor <- trimws(lr_pairs$Receptor)
lr_pairs <- unique(lr_pairs)

# Cleaning
lr_pairs <- na.omit(lr_pairs) # Remove NAs

# Remove parentheses strings
lr_pairs$Ligand <- gsub(r"{\s*\([^\)]+\)}","",as.character(lr_pairs$Ligand))
lr_pairs$Receptor <- gsub(r"{\s*\([^\)]+\)}","",as.character(lr_pairs$Receptor))

# Convert human-style format (all caps) to mouse-style format

fix_mouse_case <- function(x) {
  is_allcaps <- x == toupper(x) & nchar(x) > 1
  x[is_allcaps] <- paste0(
    toupper(substr(x[is_allcaps], 1, 1)),
    tolower(substr(x[is_allcaps], 2, nchar(x[is_allcaps])))
  )
  x
}

lr_pairs$Ligand <- fix_mouse_case(lr_pairs$Ligand)
lr_pairs$Receptor <- fix_mouse_case(lr_pairs$Receptor)

# Manual alias correction
alias_fix <- c(EPCR = "Procr")

apply_alias <- function(x) ifelse(x %in% names(alias_fix), alias_fix[x], x)

lr_pairs$Ligand   <- apply_alias(lr_pairs$Ligand)
lr_pairs$Receptor <- apply_alias(lr_pairs$Receptor)

# Sort "Integrins" issue
generic_integrin_rows <- lr_pairs$Receptor == "Integrins"
if (any(generic_integrin_rows)) {
  message(
    "Dropping generic 'Integrins' placeholder rows (Itgb1 not measured, ",
    "so no specific heterodimer is resolvable):\n",
    paste(sprintf(" %s -> Integrins", lr_pairs$Ligand[generic_integrin_rows]), collapse = "\n")
  )
}

lr_pairs <- lr_pairs[!generic_integrin_rows, ]

or_semantics_row <- lr_pairs$Receptor == "Integrins/LRP1"
lr_pairs$Receptor[or_semantics_row] <- "Lrp1"
if (any(or_semantics_row)) {
  message("Resolved 'Integrins/LRP1' (OR semantics) to Lrp1 only (integrins unmeasured).")
}

# Sort "PDGFs" issue
generic_pdgf_rows <- lr_pairs$Ligand == "PDGFs"
if (any(generic_pdgf_rows)) {
  message(
    "Dropping generic 'PDGFs' placeholder rows (no PDGF ligand gene measured):\n",
    paste(sprintf("  PDGFs -> %s", lr_pairs$Receptor[generic_pdgf_rows]), collapse = "\n")
  )
}
lr_pairs <- lr_pairs[!generic_pdgf_rows, ]

# Sort FGFR1/KLB Issue
klb_row <- lr_pairs$Receptor == "Fgfr1/klb"
lr_pairs$Receptor[klb_row] <- "Fgfr1"
if (any(klb_row)) message("Resolved 'FGFR1/KLB' to Fgfr1 only (KLB co-receptor not measured -- interaction is approximate).")

# Sort "CCL5 ligands" / "Cxcr6" issue

# Drop pairs NOT in Xenium gene panel
not_measured <- c(
  "Pdgfa", "Pdgfb", "Pdgfc", "Pdgfd",  # PDGF ligands not in panel
  "Cxcr4", "Ccr9",                      # receptors flagged absent
  "Itga1", "Itga2", "Itgb1",            # collagen-integrin subunits not measured
  "Procr",                                # EPCR/Procr absent"
  "CCL5 ligands", "Cxcr6"               # non-canonical
)

is_measured <- function(gene_str) {
  # handles "Itga4/Itgb1"-style complexes: require ALL components measured
  genes <- strsplit(gene_str, "/")[[1]]
  all(!genes %in% not_measured)
}
 
lr_pairs$keep <- mapply(function(l, r) is_measured(l) && is_measured(r),
                         lr_pairs$Ligand, lr_pairs$Receptor)
 
dropped <- lr_pairs[!lr_pairs$keep, c("Ligand", "Receptor")]
if (nrow(dropped) > 0) {
  message("Dropped (gene not measured):\n",
          paste(sprintf("  %s -> %s", dropped$Ligand, dropped$Receptor), collapse = "\n"))
}
 
lr_pairs_clean <- lr_pairs[lr_pairs$keep, c("Ligand", "Receptor")]
```

# Prepare Spatial Transcriptomics Information
```{r}
data_input <- test_obj@assays$Xenium@layers[["data"]]

rownames(data_input) <- rownames(test_obj@assays$Xenium)
colnames(data_input) <- colnames(test_obj)

meta <- data.frame(
  labels = test_obj$cell_type,
  row.names = colnames(test_obj)
)
  
spatial_locs <- data.frame(
  labels = test_obj$cell_type,
  row.names = colnames(test_obj)
)

spatial_factors <- data.frame(ratio = 1, tol = 5)

```

# Function: build custom CCDB from pair list
```{r}

build_custom_lr_db <- function(
    lr_pairs,
    species = c("mouse", "human")
) { 
  
  species <- match.arg(species)
  CellChatDB <- if (species == "mouse") CellChatDB.mouse else CellChatDB.human
  
  interaction_db <- CellChatDB$interaction
  complex_db <- CellChatDB$complex
  
  expand_to_genes <- function(name) {
    if (name %in% rownames(complex_db)) {
      subunits <- unlist(complex_db[name, ])
      subunits <- subunits[nzchar(subunits)]
      return(unique(as.character(subunits)))
    }
    name 
  }
  
  interaction_db$ligand_genes <- lapply(interaction_db$ligand, expand_to_genes)
  interaction_db$receptor_genes <- lapply(interaction_db$receptor, expand_to_genes)
  
  lr_pairs$matched <- FALSE
  matched_interaction_names <- character(0)
  
  for (i in seq_len(nrow(lr_pairs))) {
    
    lig <- lr_pairs$Ligand[i]
    rec <- lr_pairs$Receptor[i]
    
    hits <- vapply(seq_len(nrow(interaction_db)), function(j) {
      lig %in% interaction_db$ligand_genes[[j]] &&
        rec %in% interaction_db$receptor_genes[[j]]
    }, logical(1))
    
    if (any(hits)) {
      lr_pairs$matched <- TRUE
      matched_interaction_names <- union(
        matched_interaction_names,
        rownames(interaction_db)[hits]
      )
    }
  }
  
  n_matched <- sum(lr_pairs$matched)
  n_total <- nrow(lr_pairs)
  
  if (n_matched < n_total) {
    message(
      "Unmatched pairs (not found in CellChatDB, even after expanding complexes):\n",
      paste(
        sprintf(" %s -> %s", lr_pairs$Ligand[!lr_pairs$matched], lr_pairs$Receptor[!lr_pairs$matched]),
        collapse = "\n"
      ),
      "\nThese will NOT be included in the analysis unless manually added to CellChatDB "
    )
  }
  
  CellChatDB.use <- CellChatDB
  CellChatDB.use$interaction <- interaction_db[matched_interaction_names, ]
  CellChatDB.use$interaction$ligand_gens <- NULL
  CellChatDB.use$interaction$receptor_genes <- NULL
  
  list(
    CellChatDB.use = CellChatDB.use,
    match_report = lr_pairs
  )
  
  }
```



# Function: Prepare CellChat (per timepoint)
```{r}
prepare_cellchat_inputs <- function(
    seurat_obj,
    timepoint,
    time_col = "time_point",
    celltype_col = "broad_celltype",
    ratio = 1,
    tol = 5
) { 

  keep <- !is.na(seurat_obj@meta.data[[time_col]]) &
    seurat_obj@meta.data[[time_col]] == timepoint
  
  if (sum(keep) == 0) {
    stop("prepare_cellchat_inputs: no cells found for timepoint = ", timepoint)
  }
  
  sub_obj <- seurat_obj[, keep]
  
  # expression matrix
  data_input <- sub_obj@assays$Xenium@layers[["data"]]

  rownames(data_input) <- rownames(sub_obj@assays$Xenium)
  colnames(data_input) <- colnames(sub_obj)
  
  # cell type labels
  meta <- data.frame(
    labels = sub_obj@meta.data[[celltype_col]],
    row.names = colnames(sub_obj)
  )
  
  # spatial coordiantes
  spatial_locs <- data.frame(
    x = sub_obj$sdimx,
    y = sub_obj$sdimy,
    row.names = colnames(sub_obj)
  )
  
  spatial_factors <- data.frame(ratio = ratio, tol = tol)
  
  list(
    data_input = data_input,
    meta = meta,
    spatial_locs = spatial_locs,
    spatial_factors = spatial_factors,
    n_cells = sum(keep)
  )
} 
```

# Build custom LR database
```{r}
  # Build custom LR database
  
  lr_pairs <- lr_pairs_clean
  
  lr_result <- build_custom_lr_db(lr_pairs[, c("Ligand", "Receptor")], species = "mouse")
  CellChatDB.use <- lr_result$CellChatDB.use
  
  timepoints <- c("1wk", "2wk", "4wk", "12wk")
```

# Create cell chat inputs for broad type
```{r}
cellchat_inputs_broad <- setNames(
  lapply(timepoints, function(tp) {
    message("Preparing inputs for: ", tp)
    prepare_cellchat_inputs(xen1diet, timepoint = tp, celltype_col = "broad_celltype")
  }),
  timepoints
)

sapply(cellchat_inputs_broad, function(x) x$n_cells)
```

# Sanity check
```{r}
sapply(cellchat_inputs_broad, function(x) x$n_cells)
```

# Create CellChat objects
```{r}
cellchat_objects_broad <- lapply(names(cellchat_inputs_broad), function(tp) {
  inp <- cellchat_inputs_broad[[tp]]
  message("Creating CellChat object for: ", tp)
  cc <- createSpatialCellChat(
    object = inp$data_input,
    meta = inp$meta,
    group.by = "labels",
    datatype = "spatial",
    coordinates = inp$spatial_locs,
    spatial.factors = inp$spatial_factors
  )
  cc@DB <- CellChatDB.use
  cc
})
names(cellchat_objects_broad) <- names(cellchat_inputs_broad)

cellchat_objects_broad

```

# Run CellChat Inference across all timepoints
```{r}
shared_scale_distance <- 100

cellchat_objects_broad <- lapply(names(cellchat_objects_broad), function(tp) {
  
  out_file <- paste0("cellchat_", tp, "_processed.rds")
  
  # Skip re-running a timepoint that already successfully completed
  if (file.exists(out_file)) {
    message(tp, " already processed -- loading from disk instead of re-running.")
    return(readRDS(out_file))
  }
  
  message("Running inference pipeline for: ", tp)
  cc <- cellchat_objects_broad[[tp]]
  
  message("subsetData")
cc <- subsetData(cc)
saveRDS(cc, "01_subset.rds")

  message("identify genes")
cc <- identifyOverExpressedGenes(cc)
saveRDS(cc, "02_genes.rds")

  message("identify interactions")
cc <- identifyOverExpressedInteractions(cc)
saveRDS(cc, "03_interactions.rds")
  
  message("compute probabilities")
  cc <- computeCommunProb(
    cc,
    distance.use = TRUE,
    interaction.range = 250,
    scale.distance = shared_scale_distance,
    contact.dependent = TRUE,
    contact.range = 10
  )
  
  cc <- filterCommunication(cc, min.cells = 10)
  cc <- computeCommunProbPathway(cc)
  cc <- aggregateNet(cc)
  
  # Save timepoint immediately
  saveRDS(cc, out_file)
  message(tp, " cmompleted and saved [", Sys.time(), "]")
  
  cc
})
names(cellchat_objects_broad) <- names(cellchat_inputs_broad)

# final combined save
saveRDS(cellchat_objects_broad, "cellchat_objects_broad_processed.rds")

```


