---
title: "Pseudobulk Differential Expression"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

# Installs
```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("edgeR")

install.packages("RnaSeqGeneEdgeRQL")
```

# Load modules
```{r}
library(Seurat)
library(edgeR)
library(ggplot2)
library(dplyr)
library(Matrix)
library(pheatmap)
```

# Function 6: Pseudobulk ROIs
```{r}
# Usage: pseudobulked_rois <- build_pseudobulk(roi_list, add_xen1_metadata)

build_pseudobulk <- function(
    roi_list,
    metadata_fun = NULL,
    assay = "Xenium",
    layer = "counts"
) {
  
  # Build summed ROI matrix
  counts <- lapply(
  roi_list,
  function(x) {
    
    Matrix::rowSums(
      LayerData(
        x,
        assay = assay,
        layer = layer
      )
    )
  }
)

count_matrix <- do.call(
  cbind,
  counts,
)

colnames(count_matrix) <- names(roi_list)

metadata <- NULL

if (!is.null(metadata_fun)) {
  metadata <- data.frame(
    ROI = names(roi_list)
  )
  
  metadata <- metadata_fun(metadata)
  
  count_matrix <- count_matrix[
    ,
    match(metadata$ROI, colnames(count_matrix))
  ]
  
  return(list(
    counts = count_matrix,
    metadata = metadata
  ))
}
}
```

# Add metadata to matrix
```{r}
xen1_pb <- build_pseudobulk(
  xen1_rois_cortex_F,
  add_xen1_metadata
)
```

# Create DGE object
```{r}
dge <- DGEList(
  counts = xen1_pb$counts,
  samples = xen1_pb$metadata
)
```

# Filter genes
```{r}
# Filter by expression 
keep <- filterByExpr(
  dge,
  group = dge$Timepoint
)

dge <- dge[
  keep,
  ,
  keep.lib.sizes = FALSE
]
```

# Normalise
```{r}
dge <- calcNormFactors(dge)

```

# Design matrix
```{r}
design <- model.matrix(
  ~ dge$Timepoint,
  data = pb$metadata
)
design
```

# Filter genes
keep <- filterByExpr(
  dge,
  group = metadata$Timepoint
)

dge <- dge[
  keep,
  ,
  keep.lib.sizes = FALSE
]

# Normalise
dge <- calcNormFactors(dge)

# Design matrix
design <- model.matrix(
  ~ Timepoint,
  data = xen1_pb$metadata
)

# Estimate dispersal
```{r}
dge <- estimateDisp(
  dge,
  design,
  robust = TRUE
)
```

# Fit model
```{r}
fit <- glmQLFit(
  dge,
  design,
  robust = TRUE
)
```

# Quasi-likelihood F-test
```{r}
qlf <- list()
for(i in 1:ntimep){
  qlf[[i]] <- glmQLFTest(fit, contrast=contr[,i])
  qlf[[i]]$comparison <- paste0("Timepoint", levels(Timepoint)[i], "_vs_others")
}

length(qlf)
```

# Top differentially expressed genes across timepoints of healthy female cortex
```{r}
topTags(qlf[[1]], n=10L)
```

```{r}
dt <- lapply(lapply(qlf, decideTests), summary)
dt.all <- do.call("cbind", dt)
dt.all
```




# Pseudobulking with ROIs
```{r}
roi_counts <- lapply(
  xen1_rois_cortex_F,
  function(x) {
    
    Matrix::rowSums(
      LayerData(
        x,
        assay = "Xenium",
        layer = "counts"
      )
    )
  }
)

count_matrix <- do.call(
  cbind,
  roi_counts
)
```




```{r}
lcpm <- edgeR::cpm(dge, log=TRUE)
annot <- data.frame(Timepoint=paste0("Timepoint ", Timepoint))
rownames(annot) <- colnames(dge)
ann_colors <- list(cluster=2:8)
names(ann_colors$cluster) <- paste0("Timepoint ", levels(Timepoint))
pheatmap::pheatmap(lcpm[topMarkers, ], breaks=seq(-2,2,length.out=101),
                   color=colorRampPalette(c("blue","white","red"))(100), scale="row",
                   cluster_cols=TRUE, border_color="NA", fontsize_row=5,
                   treeheight_row=70, treeheight_col=70, cutree_cols=7,
                   clustering_method="ward.D2", show_colnames=FALSE,
                   annotation_col=annot, annotation_colors=ann_colors)


```





