---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---
# Gene Detection Frequencies and Spearman's Correlation
# For unclean Xen1, XeniumClean-processed Xen1, and
# the matched scRNA-seq reference. 

# Load datasets
```{r}
# original
xen1diet <- readRDS("data/xen1diet.rds")

# reference
sc_rna_seq <- readRDS("data/yuehan_stromal_integrated.rds")

# cleaned
xenium_clean <- readRDS("results/objects/xenium_clean_banksy_res1.5.rds")
```

# Same mapping for broad cell types
```{r}
sc_rna_seq$scRNAseq_broad_type <- assign_broad_cell_types(
  sc_rna_seq$cell_type
)

xen1diet$scRNAseq_broad_type <- assign_broad_cell_types(
  xen1diet$cell_type
)

xenium_clean$scRNAseq_broad_type <- assign_broad_cell_types(
  xenium_clean$cell_type
)
```

# Custom function for determining gene detection + exporting .csv
```{r}
export_gene_detection <- function(
    seurat_obj,
    cell_type_column,
    assay,
    genes,
    output_dir,
    filename_prefix
) {

  # Get raw counts and normalised expression
  counts <- GetAssayData(
    seurat_obj,
    assay = assay,
    layer = "counts"
  )

  expression <- GetAssayData(
    seurat_obj,
    assay = assay,
    layer = "data"
  )

  dir.create(
    output_dir,
    recursive = TRUE,
    showWarnings = FALSE
  )

  cell_types <- sort(
    unique(seurat_obj[[cell_type_column]][, 1])
  )

  for (cell_type in cell_types) {

    if (is.na(cell_type)) next

    message("Processing: ", cell_type)

    cells <- rownames(seurat_obj[[]])[
      seurat_obj[[cell_type_column]][, 1] == cell_type
    ]

    cells <- intersect(
      cells,
      colnames(counts)
    )

    panel_genes <- intersect(
      genes,
      rownames(counts)
    )

    if (length(cells) == 0 || length(panel_genes) == 0) {
      warning("Skipping ", cell_type)
      next
    }

    counts_sub <- counts[
      panel_genes,
      cells,
      drop = FALSE
    ]

    expression_sub <- expression[
      panel_genes,
      cells,
      drop = FALSE
    ]

    # Detection
    detected_cells <- Matrix::rowSums(
      counts_sub > 0
    )

    total_cells <- ncol(counts_sub)

    detection_frequency <-
      detected_cells / total_cells

    # Mean normalised expression
    average_expression <-
      Matrix::rowMeans(expression_sub)

    results <- data.frame(
      Cell_Type = cell_type,
      Gene = names(detected_cells),
      Detected_Cells = as.integer(detected_cells),
      Total_Cells = total_cells,
      Detection_Frequency =
        as.numeric(detection_frequency),
      Detection_Percent =
        as.numeric(detection_frequency * 100),
      Mean_Normalised_Expression =
        as.numeric(average_expression)
    )

    results$Category <- dplyr::case_when(
      results$Detection_Frequency >= 0.10 ~
        "Expressed",
      results$Detection_Frequency < 0.05 ~
        "Not Expressed",
      TRUE ~
        "Intermediate"
    )

    results <- results[
      order(-results$Detection_Frequency),
      ,
      drop = FALSE
    ]

    safe_name <- gsub(
      "[^A-Za-z0-9_]+",
      "_",
      cell_type
    )

    write.csv(
      results,
      file.path(
        output_dir,
        paste0(
          filename_prefix,
          safe_name,
          "_gene_detection.csv"
        )
      ),
      row.names = FALSE
    )
  }
}
```

# Find shared genes between all 3 datasets
```{r}
library(AnnotationDbi)
library(org.Mm.eg.db)

gene_map <- AnnotationDbi::select(
  org.Mm.eg.db,
  keys = rownames(sc_rna_seq[["RNA"]]),
  keytype = "ENSEMBL",
  columns = "SYMBOL"
)


xen1_genes <- rownames(xen1diet[["Xenium"]])
clean_genes <- rownames(xenium_clean[["XeniumClean"]])

scRNA_genes <- unique(
  gene_map$SYMBOL[!is.na(gene_map$SYMBOL)]
)

shared_genes <- Reduce(
  intersect,
  list(
    scRNA_genes,
    xen1_genes,
    clean_genes
  )
)

length(shared_genes)

scRNA_shared_genes <- gene_map$ENSEMBL[
  gene_map$SYMBOL %in% shared_genes
]

length(scRNA_shared_genes)
```

```{r}
# scRNA-seq broad 
export_gene_detection(
  seurat_obj = sc_rna_seq,
  cell_type_column = "scRNAseq_broad_type",
  assay = "RNA",
  genes = scRNA_shared_genes,
  output_dir =
    "results/Xenium_clean/Xen1_scRNAseq_gene_detection_broad_celltypes_CommonMapping",
  filename_prefix = "Xen1_scRNA-seq_"
)

# Xen1diet broad
export_gene_detection(
  seurat_obj = xen1diet,
  cell_type_column = "scRNAseq_broad_type",
  assay = "Xenium",
  genes = shared_genes,
  output_dir =
    "results/Xenium_clean/Clean_Xen1diet_gene_detection_broad_celltypes_CommonMapping",
  filename_prefix = "Xen1_"
)

# Xenium_clean broad
export_gene_detection(
  seurat_obj = xenium_clean,
  cell_type_column = "scRNAseq_broad_type",
  assay = "XeniumClean",
  genes = shared_genes,
  output_dir =
    "results/Xenium_clean/Clean_Xen1_gene_detection_broad_celltypes_CommonMapping",
  filename_prefix = "Xen1_Xenium_Clean_"
)
```

```{r}
shared_broad_types <- c(
  "Collecting duct",
  "DCT",
  "Endothelium",
  "Glomerular",
  "Immune",
  "PT",
  "Stroma",
  "TAL",
  "Thin limb",
  "Urothelium"
)

scRNA_dir <- "results/Xenium_clean/Xen1_scRNAseq_gene_detection_broad_celltypes_CommonMapping"
xen1_dir <- "results/Xenium_clean/Clean_Xen1diet_gene_detection_broad_celltypes_CommonMapping"
clean_dir <- "results/Xenium_clean/Clean_Xen1_gene_detection_broad_celltypes_CommonMapping"
```

```{r}
read_detection_files <- function(
    directory,
    files,
    dataset_name
) {

  purrr::map_dfr(files, function(file) {

    df <- read.csv(
      file.path(directory, file),
      stringsAsFactors = FALSE
    )

    df$Dataset <- dataset_name

    df

  })
}
```

```{r}
scRNA_files <- list.files(
  scRNA_dir,
  pattern = "\\.csv$",
  full.names = FALSE
)

xen1_files <- list.files(
  xen1_dir,
  pattern = "\\.csv$",
  full.names = FALSE
)

clean_files <- list.files(
  clean_dir,
  pattern = "\\.csv$",
  full.names = FALSE
)
```

```{r}
scRNA_files <- c(
  "Xen1_scRNA-seq_Collecting_duct_gene_detection.csv",
  "Xen1_scRNA-seq_DCT_gene_detection.csv",
  "Xen1_scRNA-seq_Endothelium_gene_detection.csv",
  "Xen1_scRNA-seq_Glomerular_gene_detection.csv",
  "Xen1_scRNA-seq_Immune_gene_detection.csv",
  "Xen1_scRNA-seq_PT_gene_detection.csv",
  "Xen1_scRNA-seq_Stroma_gene_detection.csv",
  "Xen1_scRNA-seq_TAL_gene_detection.csv",
  "Xen1_scRNA-seq_Thin_limb_gene_detection.csv",
  "Xen1_scRNA-seq_Urothelium_gene_detection.csv"
)

xen1_files <- c(
  "Xen1_Collecting_duct_gene_detection.csv",
  "Xen1_DCT_gene_detection.csv",
  "Xen1_Endothelium_gene_detection.csv",
  "Xen1_Glomerular_gene_detection.csv",
  "Xen1_Immune_gene_detection.csv",
  "Xen1_PT_gene_detection.csv",
  "Xen1_Stroma_gene_detection.csv",
  "Xen1_TAL_gene_detection.csv",
  "Xen1_Thin_limb_gene_detection.csv",
  "Xen1_Urothelium_gene_detection.csv"
)

clean_files <- c(
  "Xen1_Xenium_Clean_Collecting_duct_gene_detection.csv",
  "Xen1_Xenium_Clean_DCT_gene_detection.csv",
  "Xen1_Xenium_Clean_Endothelium_gene_detection.csv",
  "Xen1_Xenium_Clean_Glomerular_gene_detection.csv",
  "Xen1_Xenium_Clean_Immune_gene_detection.csv",
  "Xen1_Xenium_Clean_PT_gene_detection.csv",
  "Xen1_Xenium_Clean_Stroma_gene_detection.csv",
  "Xen1_Xenium_Clean_TAL_gene_detection.csv",
  "Xen1_Xenium_Clean_Thin_limb_gene_detection.csv",
  "Xen1_Xenium_Clean_Urothelium_gene_detection.csv"
)
```

```{r}
scRNA_detection <- read_detection_files(
  scRNA_dir,
  scRNA_files,
  "scRNA_reference"
)

xen1_detection <- read_detection_files(
  xen1_dir,
  xen1_files,
  "Xen1"
)

clean_detection <- read_detection_files(
  clean_dir,
  clean_files,
  "XeniumClean"
)
```

```{r}
scRNA_detection$GeneSymbol <- shared_gene_map$SYMBOL[
  match(
    scRNA_detection$Gene,
    shared_gene_map$ENSEMBL
  )
]

scRNA_detection$Gene <- scRNA_detection$GeneSymbol
scRNA_detection$GeneSymbol <- NULL
```

```{r}
all_detection <- dplyr::bind_rows(
  scRNA_detection,
  xen1_detection,
  clean_detection
)
```

# Make Comparison Table
```{r}
library(dplyr)
library(tidyr)

detection_comparison <- all_detection %>%
  dplyr::select(
    Cell_Type,
    Gene,
    Dataset,
    Detection_Percent
  ) %>%
  tidyr::pivot_wider(
    names_from = Dataset,
    values_from = Detection_Percent
  )
```

# Change in error table, cell type summary table, and cell type improvement table
```{r}
detection_comparison <- detection_comparison %>%
  dplyr::mutate(
    Before_Error = abs(
      Xen1 - scRNA_reference
    ),
    After_Error = abs(
      XeniumClean - scRNA_reference
    ),
    Change_in_Error = After_Error - Before_Error
  )

cell_type_summary <- detection_comparison %>%
  dplyr::group_by(Cell_Type) %>%
  dplyr::summarise(
    Median_Before_Error = median(
      Before_Error,
      na.rm = TRUE
    ),
    Median_After_Error = median(
      After_Error,
      na.rm = TRUE
    ),
    Median_Change_in_Error = median(
      Change_in_Error,
      na.rm = TRUE
    ),
    Mean_Before_Error = mean(
      Before_Error,
      na.rm = TRUE
    ),
    Mean_After_Error = mean(
      After_Error,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

cell_type_summary <- detection_comparison %>%
  dplyr::group_by(Cell_Type) %>%
  dplyr::summarise(
    Median_Before_Error = median(Before_Error, na.rm = TRUE),
    Median_After_Error = median(After_Error, na.rm = TRUE),
    Mean_Before_Error = mean(Before_Error, na.rm = TRUE),
    Mean_After_Error = mean(After_Error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    Median_Error_Reduction =
      Median_Before_Error - Median_After_Error,
    Mean_Error_Reduction =
      Mean_Before_Error - Mean_After_Error
  )

cell_type_improvement <- detection_comparison %>%
  dplyr::group_by(Cell_Type) %>%
  dplyr::summarise(
    Genes_Closer = sum(Change_in_Error < 0, na.rm = TRUE),
    Genes_Unchanged = sum(Change_in_Error == 0, na.rm = TRUE),
    Genes_Further = sum(Change_in_Error > 0, na.rm = TRUE),
    Total_Genes = dplyr::n(),
    Percent_Closer =
      Genes_Closer / Total_Genes * 100,
    Percent_Further =
      Genes_Further / Total_Genes * 100,
    .groups = "drop"
  )
```

# Make a nice table
```{r}
plot_data <- detection_comparison %>%
  dplyr::select(
    Cell_Type,
    Gene,
    Before_Error,
    After_Error
  ) %>%
  tidyr::pivot_longer(
    cols = c(Before_Error, After_Error),
    names_to = "Stage",
    values_to = "Error"
  ) %>%
  dplyr::mutate(
    Stage = dplyr::recode(
      Stage,
      Before_Error = "Before XeniumClean",
      After_Error = "After XeniumClean"
    )
  )

ggplot(
  plot_data,
  aes(
    x = Stage,
    y = Error
  )
) +
  geom_boxplot() +
  facet_wrap(
    ~ Cell_Type,
    ncol = 2
  ) +
  labs(
    x = NULL,
    y = "Absolute detection-frequency error (percentage points)"
  ) +
  theme_classic()
```